`r params$doc_title`

### 2023-02-01 scripts/MSc.Rmd
updating simulation study

library(dplyr)
library(purrr)
library(tidyr)
library(stringr)
library(tibble)
library(ggplot2)
library(ggdist)
library(gghalves)
library(ggthemes)
library(ggpubr)
library(scales)
#library(biomaRt)  # different script
library(kableExtra)
library(MendelianRandomization)
library(TwoSampleMR)
library(mr.raps)
library(MRMix)
library(MRPRESSO)
library(penalized)
library(conflicted)
source("MR_lasso.R")  # Slob&Burgess 2020
source("PVE_calculation_fns.R") 
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
```

@@

-   2022-11-12:
    -   8 SNPs removed after data harmonization, one SNP associated with FTO gene (none left)
    -   in new formula for proportion of variance in phenotype explained by given SNP (PVE) we need the MAF not the EAF
        - calculate: if $eaf>0.5$ -> $maf = 1-eaf$
        - need to change betas as well? +-
    -   We discussed $K=0.05$ in our last meeting for the "prevalence" of GBC after looking at [Cancer Today](https://gco.iarc.fr/today/home) 
        - If done this way we yield a 20% proportion of variance explained in PVE. Switch back to ASR used last week!

-   2022-11-15: 
    1. Check if formula PVE numeric can be used for binary traits as well. Can the $\%var(X)$ – numeric – be compared with $\%var(U)$ $\%var(Y)$ – binary?
    2. How is $\%var$ calculated in the MRbase package in the `mr_steiger` function (Steiger filtering)?
    3. Prefer new formula for $\%var(X)$ calculation
    4. FTO gene: Add SNP associated with FTO gene to analysis that was considered ambiguous. Check the beta estimate from UK Biobank (provide rsID to Felix: `rs1558902`)
    5. Try also $K_{GBC}= 0.05/100$ for $\%var(Y)$ (before used $ASR=1.2/100.000$)
    6. Update „Figure 1“ to 6 parts
    7. Also run BMI -> GSD, FTO -> GSD + BMI -> GBC, FTO->GBC: How different are the MR methods in estimating the causal estimate?
    8. Reference estimates for simulation: Wald estimate (FTO variant) + IVW total sample + conmix total sample
    9. Plot PVE from all simulated variants as boxplots on each dot of Figure 01

-   2022-12-05
    1. How did the large CI for the simulated PVEs arise? The simulated beta coefficients were drawn from given betas **AND** sample-size adjusted standard errors `*get simulated se_{sample size}*`. Therefore, for small sample sizes we yielded large variance for the beta coefficients. I changed the simulation and only used the given standard errors for our simulation scenarios in "scripts/simstudy.R".

@@

- Issues from discussion mentioned above
- Interchanged panel A and panel B in the first version of Figure 1, used the same colours in panel B and in the legends
- Added five estimates for "BMI -> GSD": Wald estimate (FTO), Wald estimate (MC4R variant), random IVW, MRMix, weighted mode

```{r load-data}
my_data <- as_tibble(read.delim("../data-boekstegers_felix/GST_BMI_association_results.txt"))
# GBC: new data 2022-11-04
GBC_dat <- as_tibble(read.delim("../data-boekstegers_felix/C23C24_BMI_association_results.txt"))
```

@@

my_data_harm <- my_data_harm %>% 
  filter(mr_keep==TRUE | SNP == "rs1558902")  # also include FTO SNP
```

@@

# create vectors to store pvals_n in
colnames_pval <- c()
for (i in 1:length(n)) {
  colnames_pval[i] <- paste0("p_value_exp_sample_", format(n[i], scientific = FALSE))
}
## initialize
se_update <- c()
pvals_df <- data.frame(matrix(ncol = length(colnames_pval), nrow =  nrow(my_data_harm)))
colnames(pvals_df) <- colnames_pval
for (i in 1:nrow(my_data_harm)) {
  # 1) Calculate more accurates SEs using the provided betas and Pvals 
  se_update[i] <- my_data_harm[["beta.bmi"]][i]/-qnorm(p=(my_data_harm[["p.value.bmi"]][i]/2),mean=0)

  # 2) Calculate the approximate SEs for decreasing samples sizes
  ### approximation of sd (single value)
  sd <- se_update[i] * sqrt(n.orig)
  #### approximation of standard error simulating different sample size
  se_n <- sd/(sqrt(n))

  # 3) Calculate the probability values considering the estimated betas and the corresponding SEs
  for (j in 1:length(n)) {
    pvals_n <- 2*(1-pnorm(abs(my_data_harm[["beta.bmi"]][i])/abs(se_n[j]), mean=0, 
                          lower.tail = TRUE))
    pvals_df[i, j] <- pvals_n
  }
}

@@

# pvals_df %>%
#   filter(p_value_exp_sample_100000< 5*10^(-6)) %>% 
#   count()

my_data_harm <- as_tibble(cbind(my_data_harm, pvals_df))
```

@@

## CALCULATE explained variance
my_data_harm <- my_data_harm %>% 
  mutate(explained_variance_BMI = explained_variance_numeric(eaf = eaf.bmi, beta = beta.bmi), 
         # USE THIS !
         explained_variance_BMI2 = explained_variance_numeric2(maf=eaf.bmi, 
                                                           beta=beta.bmi, 
                                                           se_beta=se.bmi, 
                                                           samplesize=n[length(n)]

         ),
         explained_variance_GSD = explained_variance_binary(PA = eaf.gsd, 
                                                            RR1 = exp(beta.gsd), 
                                                            RR2 = 1+(2*(exp(beta.gsd)-1)), 
                                                            K = 0.1  # assumed prevalence in population
                                                            )$Vg,
         explained_variance_GBC = explained_variance_binary(PA = eaf.gbc, 
                                                            RR1 = exp(beta.gbc), 
                                                            RR2 = 1+(2*(exp(beta.gbc)-1)), # exp(beta.gbc)^2, 
                                                            K =1.2/100000  # assumed prevalence in population
                                                            )$Vg

         )
# sum(my_data_harm$explained_variance_GBC)
```

@@

filenamefig00_a <- paste0("../output/plots/", Sys.Date(), "_explained_var_X_old_formula.png")
ggsave(filename = filenamefig00_a, plot = fig00, 
       width = 10, height = 6, 
       units = "in"  # default
       )

@@

for (i in 1:length(colnames_pval) ) {
  threshold <- 5*10^(-6)
  select_col <- colnames_pval[i]
  # calculate cum explained variance and count of IVs
  summarydat <- my_data_harm %>% 
    filter(!!sym(select_col) < threshold) %>% 
    select(explained_variance_BMI, explained_variance_BMI2, explained_variance_GSD, explained_variance_GBC) %>% 
    # cumulative PVE
    summarise(expl_variances =across(.cols = everything(), function(x) sum(x, na.rm = TRUE)),
              count_IVs = n())
  # fill up df
  df_cum_expl_var[i, 2] <- summarydat$expl_variances[1]
  df_cum_expl_var[i, 3] <- summarydat$expl_variances[2]
  df_cum_expl_var[i, 4] <- summarydat$expl_variances[3]
  df_cum_expl_var[i, 5] <- summarydat$expl_variances[4]
  df_cum_expl_var[i, 6] <- summarydat$count_IVs
}
```

@@

df_cum_expl_var <- df_cum_expl_var %>% 
  mutate(
    cum_expl_varX = cum_expl_varX *100, # in percent 
    cum_expl_varX2 = cum_expl_varX2 *100,
    cum_expl_varU = cum_expl_varU *100,
    cum_expl_varY = cum_expl_varY *100,
    count_IVs_scaled = count_IVs*scale_factor)

plot_data1 <- df_cum_expl_var %>% 
  pivot_longer(cols = c(cum_expl_varX, cum_expl_varU, cum_expl_varY)) %>% 
  mutate(name = factor(name, labels = c("% var(U)", "% var(X)", "% var(Y)")),
         name = factor(name, levels = c("% var(X)", "% var(Y)", "% var(U)") )
         )

N_annotation <- c(10000, 100000, 200000, 300000)
## plot
ggplot(plot_data1, aes(x=n, y=value, col=name))+
  geom_point(alpha=0.6)+
  geom_line(alpha=0.6)+
  geom_line(aes(x=n, y=count_IVs_scaled, col="Number of IVs"), linetype="dotdash")+
  scale_x_continuous(name = paste0("Size of sample ", as.roman(1)),  # 2-sample-MR or label "Study Size"
                     labels = comma, 
                     breaks = N_annotation, #c( 10000, n[7:length(n)]),
                     limits = c(10000, 300000))+
  #ylab("% variance explained by IVs")+
  labs(title = "Figure 1",
       col='', 
       caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  scale_y_continuous(
    # Features of the first axis
    name = "% variance explained by IVs",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./scale_factor, name="Number of IVs")
  ) + 
  theme_few()+
  #scale_colour_few() +
  #scale_color_manual(name = "", values = c("% var(U)" = "red", 
  #                                         "% var(X)" = "blue", 
  #                                         "% var(Y)" = "darkblue", 
  #                                         "Number of IVs" = "black"
  #                                         ))
  #
  scale_color_manual(name = "", values = c("% var(U)" = ggthemes_data$few$colors$Dark[2,2][[1]], 
                                           "% var(X)" = ggthemes_data$few$colors$Dark[3,2][[1]], 
                                           "% var(Y)" = ggthemes_data$few$colors$Dark[4,2][[1]], 
                                           "Number of IVs" = "black"),
                     # order
                     breaks = c( "Number of IVs", "% var(X)", "% var(Y)", "% var(U)")
  ) -> fig01.update

# fig01.update   # old formula for PVE exposure X (without sample size adjusting)
# 
# scale_color_manual(name = "Y series", values = c("Y1" = "darkblue", "Y2" = "red"))

@@

# SAVE
filenamefig01_a <- paste0("../output/plots/", Sys.Date(), "_figure01_old_PVE_X_formula.png")
filenamefig01_b <- paste0("../output/plots/", Sys.Date(), "_figure01_new_PVE_X_formula.png")
ggsave(filename = filenamefig01_a, plot = fig01.update, 
       width = 10, height = 6, 
       units = "in"  # default
       )

ggsave(filename = filenamefig01_b, plot = fig01b.update, 
       width = 10, height = 6, 
       units = "in"  # default
       )
```

@@

```{r mafVSbetas}
plot_data_2 <- my_data_harm %>% 
  mutate(maf.bmi = ifelse(eaf.bmi >0.5, 1-eaf.bmi, eaf.bmi)) %>% 
  select(maf.bmi, beta.bmi)

ggplot(data = plot_data_2, aes(x=maf.bmi, y=abs(beta.bmi)))+
  geom_point(alpha=0.6)+
  geom_hline(yintercept=median(abs(plot_data_2$beta.bmi)), 
             col=ggthemes_data$few$colors$Dark[6,2][[1]], alpha=0.6, linetype=2)+
  geom_vline(xintercept=median(plot_data_2$maf.bmi), 
             col=ggthemes_data$few$colors$Dark[6,2][[1]], alpha=0.6, linetype=2)+
  annotate(geom = "text", x = 0.15, y = 0.045, 
           label="relatively rare\nrelatively strong", 
           color= ggthemes_data$few$colors$Dark[2,2][[1]]
             )+
  annotate(geom = "text", x = 0.42, y = 0.05, 
           label="relatively common\nrelatively strong", 
           color=ggthemes_data$few$colors$Dark[3,2][[1]])+
  annotate(geom = "text", x = 0.1, y = 0.015, 
           label="relatively rare\nrelatively weak", color=ggthemes_data$few$colors$Dark[4,2][[1]])+
  annotate("segment", x=0.44,xend=0.42,y=0.025,yend=0.013,arrow=arrow(), 
           color=ggthemes_data$few$colors$Dark[5,2][[1]])+
  annotate(geom = "text", x = 0.44, y = 0.03, 
           label="relatively common\nrelatively weak", 
           color=ggthemes_data$few$colors$Dark[5,2][[1]])+
  #geom_text(x=0.1, y=0.05, label="hello", family = "Times New Roman",
  #          size=6, col="red")+
  xlab("Minor allele frequency")+
  ylab("Exposure effect size")+  # abs(beta_BMI)
  labs(title = "",
       subtitle = "Define different genetic architectures",
       #col='', 
       #caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  theme_few()+
  scale_colour_few() -> fig02.update
fig02.update

@@

```{r figure1quadrants, fig.show="hold", out.width="50%", warning=FALSE}

plotfigure01 <- function(data, effect_size="strong", maf="rare"){
  ## data: my_data_harm
  # effect size: "strong" or "weak"
  # maf: "rare" or "common
  df_cum_expl_var <- tibble(n=n, 
                            cum_expl_varX = rep(NA, length(n)), 
                            cum_expl_varX2 = rep(NA, length(n)), 
                            cum_expl_varU = rep(NA, length(n)), 
                            cum_expl_varY = rep(NA, length(n)),
                            count_IVs = rep(NA, length(n))
  )
  # create maf col
  data <- data %>% 
      mutate(maf.bmi = ifelse(eaf.bmi >0.5, 1-eaf.bmi, eaf.bmi))

  ## filter for quadrant
  if (effect_size=="strong" & maf == "rare") {
    data <- data %>% 
      filter(abs(beta.bmi) > median(abs(beta.bmi)) &  maf.bmi < median(maf.bmi))
  } else if (effect_size=="strong" & maf == "common") {
    data <- data %>% 
      filter(abs(beta.bmi) > median(abs(beta.bmi)) &  maf.bmi > median(maf.bmi))
  } else if (effect_size=="weak" & maf == "common") {
    data <- data %>% 
      filter(abs(beta.bmi) < median(abs(beta.bmi)) &  maf.bmi > median(maf.bmi))
  } else if (effect_size=="weak" & maf == "rare") {
    data <- data %>% 
      filter(abs(beta.bmi) < median(abs(beta.bmi)) &  maf.bmi < median(maf.bmi))
  }

  # create data for plot
  for (i in 1:length(colnames_pval) ) {
    threshold <- 5*10^(-6)
    select_col <- colnames_pval[i]
    # calculate cum explained variance and count of IVs
    summarydat <- data %>% 
      filter(!!sym(select_col) < threshold) %>% 
      select(explained_variance_BMI, explained_variance_BMI2, explained_variance_GSD, explained_variance_GBC) %>% 
      summarise(expl_variances =across(.cols = everything(), function(x) sum(x)),
                count_IVs = n())
    # fill up df
    df_cum_expl_var[i, 2] <- summarydat$expl_variances[1]
    df_cum_expl_var[i, 3] <- summarydat$expl_variances[2]
    df_cum_expl_var[i, 4] <- summarydat$expl_variances[3]
    df_cum_expl_var[i, 5] <- summarydat$expl_variances[4]
    df_cum_expl_var[i, 6] <- summarydat$count_IVs
  }

  ## plot fig 01 
  scale_factor <- 4.5/284
  df_cum_expl_var <- df_cum_expl_var %>% 
    mutate(
      cum_expl_varX = cum_expl_varX *100, # in percent 
      cum_expl_varX2 = cum_expl_varX2 *100,
      cum_expl_varU = cum_expl_varU *100,
      cum_expl_varY = cum_expl_varY *100,
      count_IVs_scaled = count_IVs*scale_factor)

  plot_data1 <- df_cum_expl_var %>% 
    pivot_longer(cols = c(cum_expl_varX, cum_expl_varU, cum_expl_varY)) %>% 
    mutate(name = factor(name, labels = c("% var(U)", "% var(X)", "% var(Y)")),
           name = factor(name, levels = c("% var(X)", "% var(Y)", "% var(U)") )
           )

  N_annotation <- c(10000, 100000, 200000, 300000)
  ## plot
  plotsubtitle <- paste0("relatively ", maf ,", relatively ", effect_size)
  ggplot(plot_data1, aes(x=n, y=value, col=name))+
    geom_point(alpha=0.6)+
    geom_line(alpha=0.6)+
    geom_line(aes(x=n, y=count_IVs_scaled, col="Number of IVs"), linetype="dotdash")+
    scale_x_continuous(name = paste0("Size of sample ", as.roman(1)),  # 2-sample-MR or label "Study Size"
                       labels = comma, 
                       breaks = N_annotation, #c( 10000, n[7:length(n)]),
                       limits = c(10000, 300000))+
    #ylab("% variance explained by IVs")+
    labs(title = "", subtitle = plotsubtitle,
         col='', 
         caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits   explained by the IVs"
         ) +
    scale_y_continuous(
      # Features of the first axis
      name = "% variance explained by IVs",
      # Add a second axis and specify its features
      sec.axis = sec_axis(~./scale_factor, name="Number of IVs"),
      limits = c(0,2)
    ) + 
    theme_few()+
    #scale_colour_few() +
    #scale_color_manual(name = "", values = c("% var(U)" = "red", 
    #                                         "% var(X)" = "blue", 
    #                                         "% var(Y)" = "darkblue", 
    #                                         "Number of IVs" = "black"
    #                                         ))
    #
    scale_color_manual(name = "", values = c("% var(U)" = ggthemes_data$few$colors$Dark[2,2][[1]], 
                                             "% var(X)" = ggthemes_data$few$colors$Dark[3,2][[1]], 
                                             "% var(Y)" = ggthemes_data$few$colors$Dark[4,2][[1]], 
                                             "Number of IVs" = "black"),
                       # order
                       breaks = c( "Number of IVs", "% var(X)", "% var(Y)", "% var(U)")
    ) -> figure01.quadrant
  print(figure01.quadrant)

  filenamefig01_quadrant <- paste0("../output/plots/", Sys.Date(), "_figure01_", 
                                   "EFFECT_SIZE_", effect_size,
                                   "_VARIANTS_", maf,
                                   ".png")
  ggsave(filename = filenamefig01_quadrant, plot = figure01.quadrant, 
         width = 10, height = 6, 
         units = "in"  # default
         )

  # store plot in global environment
  plotname <<- paste0("figure01_EFFECT_SIZE_", effect_size, "_VARIANTS_", maf)
  assign(plotname, figure01.quadrant, envir = .GlobalEnv)
  #list2env(assign(plotname, figure01.quadrant), envir = .GlobalEnv)

# create all combis
effect_size <- c("strong", "weak")
maf <- c("rare", "common")
all_quadrants <- tidyr::crossing(effect_size, maf)

####### PLOT NOW  ---
for (i in 1:nrow(all_quadrants)) {
  plotfigure01(data = my_data_harm, 
               effect_size = all_quadrants$effect_size[i], 
               maf = all_quadrants$maf[i])
}

@@

## load results
est.bmi.gsd <- readRDS(file = dplyr::last(list.files("../output/Rdata/", pattern = "_est_BMI_GSD.rds", full.names = TRUE)))

# biology plausibility FTO & MC4R
# ## The mr_ivw function implements the inverse-variance method, informally 
# known as the "Toby Johnson" method. With a single genetic variant, this 
# is simply the ratio method.
fto_wald <- mr_wald_ratio(
  my_data_harm$beta.bmi[my_data_harm$SNP == "rs1558902"], 
  my_data_harm$beta.gsd[my_data_harm$SNP == "rs1558902"], 
  my_data_harm$se.bmi[my_data_harm$SNP == "rs1558902"], 
  my_data_harm$se.gsd[my_data_harm$SNP == "rs1558902"]
)
mc4r_wald <- mr_wald_ratio(
  my_data_harm$beta.bmi[my_data_harm$SNP == "rs6567160"], 
  my_data_harm$beta.gsd[my_data_harm$SNP == "rs6567160"], 
  my_data_harm$se.bmi[my_data_harm$SNP == "rs6567160"], 
  my_data_harm$se.gsd[my_data_harm$SNP == "rs6567160"]
)
# add rows
est.bmi.gsd <- est.bmi.gsd %>% 
  add_row(mr_methods = "FTO_Wald", estimate = fto_wald$b, se.estimate=fto_wald$se) %>% 
  add_row(mr_methods = "MC4R_Wald", estimate = mc4r_wald$b, se.estimate=mc4r_wald$se)

@@

res <- TwoSampleMR::mr(harmonised_dat4TwoSampleMR)
res <- res %>% mutate(OR = exp(b), CI.lower= exp(b-se), CI.upper=exp(b+se))

@@

est %>% 
  mutate(OR=exp(estimate), CI.lower=exp(estimate-se.estimate),CI.upper=exp(estimate+se.estimate)) %>%  
  kable(digits=3, caption = "MR Results for GBC (outcome) using genetic variants as instrumental variables for BMI") %>% 
  kable_classic(full_width = FALSE, html_font = "Cambria")
```

@@

```{r simulated-pves, fig.height=6, fig.width=10, warning=FALSE}
# retrieve PVE colnames
colname_pve_X <-  colnames(est_sim)[str_detect(colnames(est_sim), pattern = "pve_X_sim")]
colname_pve_U <-  colnames(est_sim)[str_detect(colnames(est_sim), pattern = "pve_U_sim")]
colname_pve_Y <-  colnames(est_sim)[str_detect(colnames(est_sim), pattern = "pve_Y_sim")]

>>>>

library(dplyr)
library(purrr)
library(tidyr)
library(stringr)
library(tibble)
library(ggplot2)
library(ggdist)
library(gghalves)
library(ggthemes)
library(ggpubr)
library(scales)
#library(biomaRt)  # different script
library(kableExtra)
library(MendelianRandomization)
library(TwoSampleMR)
library(mr.raps)
library(MRMix)
library(MRPRESSO)
library(penalized)
library(conflicted)
source("helper/MR_lasso.R")  # Slob&Burgess 2020
source("helper/PVE_calculation_fns.R") 
source("helper/plotfigure01fn.R") 
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
```

@@

-   NA

@@

-   All genetic associations with BMI > 0 (for association statistics < 0, sign of the association statistic with GSD changed respectively)

```{r load-data}
my_data <- as_tibble(read.delim("../data-boekstegers_felix/GST_BMI_association_results.txt"))
# GBC: new data 2022-11-04
GBC_dat <- as_tibble(read.delim("../data-boekstegers_felix/C23C24_BMI_association_results.txt"))

# FTO data
FTO_dat <- as_tibble(read.delim("../data-boekstegers_felix/2023-01-19_FTO/FTO_association_results.txt"))
```

@@

my_data_harm <- my_data_harm %>% 
  filter(mr_keep==TRUE | SNP == "rs1558902") %>%  # also include FTO SNP 
  # change sign of beta after checking consistency of data (see mail felix 2023-01-19)
  mutate(beta.bmi = ifelse(SNP == "rs1558902", -beta.bmi, beta.bmi))

my_data_harm <- my_data_harm %>% 
  mutate(
    beta.gsd = ifelse(beta.bmi<0, -beta.gsd, beta.gsd),
    beta.gbc = ifelse(beta.bmi<0, -beta.gbc, beta.gbc),
    # change last beta.bmi, otherwise not negative any more ;)
    beta.bmi = ifelse(beta.bmi<0, -beta.bmi, beta.bmi)
  )
```

@@

# create vectors to store pvals_n in
colnames_pval <- c()
for (i in 1:length(n)) {
  colnames_pval[i] <- paste0("p_value_exp_sample_", format(n[i], scientific = FALSE))
}
## initialize
se_update <- c()
pvals_df <- data.frame(matrix(ncol = length(colnames_pval), nrow =  nrow(my_data_harm)))
colnames(pvals_df) <- colnames_pval
for (i in 1:nrow(my_data_harm)) {
  # 1) Calculate more accurates SEs using the provided betas and Pvals 
  se_update[i] <- my_data_harm[["beta.bmi"]][i]/-qnorm(p=(my_data_harm[["p.value.bmi"]][i]/2),mean=0)
  # 2) Calculate the approximate SEs for decreasing samples sizes
  ### approximation of sd (single value)
  sd <- se_update[i] * sqrt(n.orig)
  #### approximation of standard error simulating different sample size
  se_n <- sd/(sqrt(n))
  # 3) Calculate the probability values considering the estimated betas and the corresponding SEs
  for (j in 1:length(n)) {
    pvals_n <- 2*(1-pnorm(abs(my_data_harm[["beta.bmi"]][i])/abs(se_n[j]), mean=0, 
                          lower.tail = TRUE))
    pvals_df[i, j] <- pvals_n
  }
}

@@

# pvals_df %>%
#   filter(p_value_exp_sample_100000< 5*10^(-6)) %>% 
#   count()
my_data_harm <- as_tibble(cbind(my_data_harm, pvals_df))
```

@@

## CALCULATE explained variance for total sample size
my_data_harm <- my_data_harm %>%
  mutate(
    explained_variance_BMI = explained_variance_numeric(eaf = eaf.bmi, beta = beta.bmi),
    # USE THIS !
    explained_variance_BMI2 = explained_variance_numeric2(
      maf = eaf.bmi, beta = beta.bmi,
      se_beta = se.bmi,
      samplesize = n[length(n)]
    ),
    explained_variance_GSD = explained_variance_binary(
      PA = eaf.gsd,
      RR1 = exp(beta.gsd),
      RR2 = 1 + (2 * (exp(beta.gsd) - 1)),
      K = 0.1 # assumed prevalence in population
    )$Vg,
    explained_variance_GBC = explained_variance_binary(
      PA = eaf.gbc,
      RR1 = exp(beta.gbc),
      RR2 = 1 + (2 * (exp(beta.gbc) - 1)), # exp(beta.gbc)^2,
      K = 1.2 / 100000 # assumed prevalence in population
    )$Vg
  )

# sum(my_data_harm$explained_variance_GBC)
```

@@

# filenamefig00_a <- paste0("../output/plots/", Sys.Date(), "_explained_var_X_old_formula.png")
# ggsave(filename = filenamefig00_a, plot = fig00, 
#        width = 10, height = 6, 
#        units = "in"  # default
#        )

@@

threshold <- 5*10^(-6)
for (i in 1:length(colnames_pval) ) {
  select_col <- colnames_pval[i]
  # calculate cum explained variance and count of IVs
  summarydat <- my_data_harm %>% 
    filter(!!sym(select_col) < threshold) %>% 
    select(explained_variance_BMI, explained_variance_BMI2, explained_variance_GSD, explained_variance_GBC) %>% 
    # cumulative PVE
    summarise(expl_variances =across(.cols = everything(), function(x) sum(x, na.rm = TRUE)),
              count_IVs = n())
  # fill up df
  df_cum_expl_var[i, 2] <- summarydat$expl_variances[1]
  df_cum_expl_var[i, 3] <- summarydat$expl_variances[2]
  df_cum_expl_var[i, 4] <- summarydat$expl_variances[3]
  df_cum_expl_var[i, 5] <- summarydat$expl_variances[4]
  df_cum_expl_var[i, 6] <- summarydat$count_IVs
}

# for the numeric traits we have to map over sample size and recalculate se_updated
n.orig_sqrt <- sqrt(n.orig)
explained_variance_BMI2_sample_size <- map(n, ~explained_variance_numeric2(maf = my_data_harm$eaf.bmi, beta = my_data_harm$beta.bmi, 
                                                                           se_beta = abs(my_data_harm$beta.bmi / -qnorm(p=(my_data_harm$p.value.bmi / 2), mean = 0) * n.orig_sqrt / sqrt(.x)), 
                                                                           samplesize = .x))

cum_expl_varX2_sample_size <- c()
for (mysamples in 1:length(colnames_pval)) {
  temp_dat <- my_data_harm %>% 
    select(all_of(colnames_pval[mysamples])) %>% 
    mutate(explained_variance_BMI2_sample_size = explained_variance_BMI2_sample_size[[mysamples]]) %>% 
    filter(!!sym(colnames_pval[mysamples]) < threshold)
  #print(temp_dat)
  cum_expl_varX2_sample_size <- append(cum_expl_varX2_sample_size, sum(temp_dat$explained_variance_BMI2_sample_size))
}

df_cum_expl_var <- df_cum_expl_var %>% 
  cbind(cum_expl_varX2_sample_size) %>% 
  relocate(cum_expl_varX2_sample_size, .before = cum_expl_varU)

ggplot(df_cum_expl_var, aes(x=n))+
  geom_line(aes(y=cum_expl_varX2_sample_size*100), col="red")+
  geom_line(aes(y=cum_expl_varX2*100))
```

@@

df_cum_expl_var <- df_cum_expl_var %>% 
  mutate(
    cum_expl_varX = cum_expl_varX *100, # in percent 
    cum_expl_varX2 = cum_expl_varX2_sample_size *100,  #  CAVE CHANGED HERE TO INCLUDE SAMPLE SIZE DEPENDENT SE and PVE CALC!
    cum_expl_varU = cum_expl_varU *100,
    cum_expl_varY = cum_expl_varY *100,
    count_IVs_scaled = count_IVs*scale_factor)

@@

filenamefig01_b <- paste0("../output/plots/", Sys.Date(), "_figure01_new_PVE_X_formula.png")
ggsave(filename = filenamefig01_b, plot = fig01b.update, 
       width = 10, height = 6, 
       units = "in"  # default
       )
```

@@

```{r mafVSbetas}
plot_data_2 <- my_data_harm %>% 
  mutate(maf.bmi = ifelse(eaf.bmi >0.5, 1-eaf.bmi, eaf.bmi)) %>% 
  select(maf.bmi, beta.bmi, gene_label, fto_col_annot)

ggplot(data = plot_data_2, aes(x=maf.bmi, y=abs(beta.bmi), 
                               col=fto_col_annot, label=gene_label))+
  geom_point(alpha=0.6)+
  ggrepel::geom_text_repel(show.legend = FALSE)+
  geom_hline(yintercept=median(abs(plot_data_2$beta.bmi)), 
             col=ggthemes_data$few$colors$Dark[6,2][[1]], alpha=0.6, linetype=2)+
  geom_vline(xintercept=median(plot_data_2$maf.bmi), 
             col=ggthemes_data$few$colors$Dark[6,2][[1]], alpha=0.6, linetype=2)+
  annotate(geom = "text", x = 0.15, y = 0.045, 
           label="relatively rare\nrelatively strong", 
           color= ggthemes_data$few$colors$Dark[2,2][[1]]
             )+
  annotate(geom = "text", x = 0.42, y = 0.05, 
           label="relatively common\nrelatively strong", 
           color=ggthemes_data$few$colors$Dark[3,2][[1]])+
  annotate(geom = "text", x = 0.1, y = 0.015, 
           label="relatively rare\nrelatively weak", color=ggthemes_data$few$colors$Dark[4,2][[1]])+
  annotate("segment", x=0.44,xend=0.42,y=0.025,yend=0.013,arrow=arrow(), 
           color=ggthemes_data$few$colors$Dark[5,2][[1]])+
  annotate(geom = "text", x = 0.44, y = 0.03, 
           label="relatively common\nrelatively weak", 
           color=ggthemes_data$few$colors$Dark[5,2][[1]])+
  #geom_text(x=0.1, y=0.05, label="hello", family = "Times New Roman",
  #          size=6, col="red")+
  xlab("Minor allele frequency")+
  ylab("Exposure effect size")+  # abs(beta_BMI)
  labs(title = "",
       subtitle = "Define different genetic architectures",
       col='SNP', 
       #caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  theme_few()+
  scale_colour_few() -> fig02.update
fig02.update

@@

```{r figure1quadrants, fig.show="hold", out.width="50%", warning=FALSE}
## plotfigure sourced from helper functions ----------------------------------
# create all combis
effect_size <- c("strong", "weak")
maf <- c("rare", "common")
all_quadrants <- tidyr::crossing(effect_size, maf)

####### PLOT NOW  ---
for (i in 1:nrow(all_quadrants)) {
  plotfigure01(data = my_data_harm, 
               effect_size = all_quadrants$effect_size[i], 
               maf = all_quadrants$maf[i], 
               n=n
               )
}

@@

## load results from script `perform_mr.R`
est.bmi.gsd <- readRDS(file = dplyr::last(list.files("../output/Rdata/", pattern = "_est_BMI_GSD.rds", full.names = TRUE)))

# biology plausibility FTO & MC4R
# ## The mr_ivw function implements the inverse-variance method, informally 
# known as the "Toby Johnson" method. With a single genetic variant, this 
# is simply the ratio method.
fto_wald <- mr_wald_ratio(
  my_data_harm$beta.bmi[my_data_harm$SNP == "rs1558902"], 
  my_data_harm$beta.gsd[my_data_harm$SNP == "rs1558902"], 
  my_data_harm$se.bmi[my_data_harm$SNP == "rs1558902"], 
  my_data_harm$se.gsd[my_data_harm$SNP == "rs1558902"]
)

fto_wald_ukbiobank <- mr_wald_ratio(
  b_exp = exp(FTO_dat$beta[3])-1, # BMI UK Biobank log-transformed  (before Hoffmann GERA + GIANT)
  b_out = FTO_dat$beta[2], # GSD UKB as before
  se_exp = exp(FTO_dat$se[3])-1, 
  se_out = FTO_dat$se[2]
  )

fto_wald_ukbiobank_noexp <- mr_wald_ratio(
  b_exp = (FTO_dat$beta[3]), # BMI UK Biobank log-transformed  (before Hoffmann GERA + GIANT)
  b_out = FTO_dat$beta[2], # GSD UKB as before
  se_exp = (FTO_dat$se[3]), 
  se_out = FTO_dat$se[2]
  )

mc4r_wald <- mr_wald_ratio(
  my_data_harm$beta.bmi[my_data_harm$SNP == "rs6567160"], 
  my_data_harm$beta.gsd[my_data_harm$SNP == "rs6567160"], 
  my_data_harm$se.bmi[my_data_harm$SNP == "rs6567160"], 
  my_data_harm$se.gsd[my_data_harm$SNP == "rs6567160"]
)
# add rows
est.bmi.gsd <- est.bmi.gsd %>% 
  add_row(mr_methods = "FTO_Wald", estimate = fto_wald$b, se.estimate=fto_wald$se) %>% 
  add_row(mr_methods = "MC4R_Wald", estimate = mc4r_wald$b, se.estimate=mc4r_wald$se) %>% 
  arrange(desc(estimate))

@@

res <- TwoSampleMR::mr(harmonised_dat4TwoSampleMR)
res <- res %>% mutate(OR = exp(b), CI.lower= exp(b-se), CI.upper=exp(b+se)) %>% 
  arrange(desc(b))

@@

est %>% 
  mutate(OR=exp(estimate), CI.lower=exp(estimate-se.estimate),CI.upper=exp(estimate+se.estimate)) %>%  
  arrange(desc(estimate)) %>% 
  kable(digits=3, caption = "MR Results for GBC (outcome) using genetic variants as instrumental variables for BMI") %>% 
  kable_classic(full_width = FALSE, html_font = "Cambria")
```

@@

```{r simulated-pves, fig.height=6, fig.width=10, warning=FALSE}
# retrieve PVE colnames
colname_pve_X <-  colnames(est_sim)[str_detect(colnames(est_sim), pattern = "pve_X_sim")]
# PVE_Y, PVE_U removed on 2023-01-24 (see github)
